---
title: "Genetic Variant Identification and Disease Association Analysis (R)"
output: html_notebook
---

# Genetic Variant Identification and Disease Association Analysis

This notebook provides comprehensive analysis of genetic variants including:
- Single-nucleotide polymorphisms (SNPs)
- Insertions/Deletions (Indels)
- Structural variants

## Dataset
- **Source**: genomics_data.csv
- **Format**: DNA sequences with disease association labels
- **Labels**: 0 = Control, 1 = Disease

```{r setup}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(gridExtra)
library(corrplot)
library(stringr)

# Source the variant identification functions
source("../src/variant_identification.R")
source("../src/variant_analysis.R")

cat("Libraries loaded successfully!\n")
```

## 1. Load and Explore Data

```{r load-data}
# Load the genomics data
data_path <- "../data/genomics_data.csv"
df <- read.csv(data_path, stringsAsFactors = FALSE)

cat("Dataset shape:", nrow(df), "rows,", ncol(df), "columns\n")
cat("\nFirst few rows:\n")
print(head(df))
cat("\nData types:\n")
print(str(df))
cat("\nLabel distribution:\n")
print(table(df$Labels))
cat("\nSequence length:", nchar(df$Sequences[1]), "nucleotides\n")
```

## 2. Calculate Consensus Sequence

```{r consensus}
# Calculate consensus sequence from all sequences
sequences <- df$Sequences
consensus <- calculate_consensus_sequence(sequences)

cat("Consensus sequence length:", nchar(consensus), "\n")
cat("Consensus sequence (first 50 bases):", substr(consensus, 1, 50), "\n")
cat("\nNucleotide composition in consensus:\n")
nuc_counts <- table(strsplit(consensus, "")[[1]])
for (nuc in names(nuc_counts)) {
  cat("  ", nuc, ":", nuc_counts[nuc], "(", round(nuc_counts[nuc]/nchar(consensus)*100, 2), "%)\n")
}
```

## 3. Identify Variants

```{r identify-variants}
# Process all sequences and identify variants
cat("Processing sequences and identifying variants...\n")
results_df <- process_genomics_data(data_path, "../results/variant_analysis_results_R.csv")

cat("\nAnalysis complete!\n")
cat("Total sequences analyzed:", nrow(results_df), "\n")
cat("\nFirst few results:\n")
print(head(results_df[, c("sequence_id", "label", "num_snps", "num_indels", 
                          "num_structural_variants", "total_variants")], 10))
```

## 4. Summary Statistics

```{r summary}
# Generate summary statistics
summary_stats <- create_summary_statistics(results_df)
cat("Summary Statistics:\n")
print(summary_stats)
```

## 5. Disease Association Analysis

```{r disease-assoc}
# Analyze disease associations
disease_assoc <- analyze_disease_association(results_df)
cat("Disease Association Analysis:\n")
print(disease_assoc)

# Statistical test
control_variants <- results_df$total_variants[results_df$label == 0]
disease_variants <- results_df$total_variants[results_df$label == 1]

t_test <- t.test(control_variants, disease_variants)
cat("\nT-test results (Control vs Disease):\n")
cat("  T-statistic:", t_test$statistic, "\n")
cat("  P-value:", t_test$p.value, "\n")
cat("  Significant difference:", ifelse(t_test$p.value < 0.05, "Yes", "No"), "(α=0.05)\n")
```

## 6. Visualizations

```{r visualizations}
# Create visualizations
create_visualizations(results_df, "../results")

# Display some plots inline
p1 <- ggplot(results_df, aes(x = num_snps)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of SNPs per Sequence",
       x = "Number of SNPs", y = "Frequency") +
  theme_minimal()

p2 <- ggplot(results_df, aes(x = num_indels)) +
  geom_histogram(bins = 30, fill = "orange", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Indels per Sequence",
       x = "Number of Indels", y = "Frequency") +
  theme_minimal()

results_df$label_factor <- factor(results_df$label, levels = c(0, 1), 
                                  labels = c("Control", "Disease"))

p3 <- ggplot(results_df, aes(x = label_factor, y = total_variants, fill = label_factor)) +
  geom_boxplot() +
  labs(title = "Total Variants by Disease Label",
       x = "Disease Label", y = "Number of Total Variants") +
  theme_minimal() +
  theme(legend.position = "none")

# Correlation matrix
correlation_cols <- c("num_snps", "num_indels", "num_structural_variants", "total_variants", "label")
corr_matrix <- cor(results_df[, correlation_cols])

grid.arrange(p1, p2, p3, ncol = 2, nrow = 2)
```

## 7. Detailed Variant Analysis

```{r detailed}
# Examine sequences with highest variant counts
cat("Sequences with most variants:\n")
top_variants <- results_df %>% 
  arrange(desc(total_variants)) %>% 
  head(10)
print(top_variants[, c("sequence_id", "label", "num_snps", "num_indels", 
                      "num_structural_variants", "total_variants")])

# Example: Detailed analysis of first sequence with variants
example_idx <- which(results_df$total_variants > 0)[1]
example_seq <- df$Sequences[example_idx + 1]  # +1 because sequence_id is 0-indexed
example_analysis <- analyze_sequence(example_seq, consensus)

cat("\n\nExample Analysis (Sequence ID:", example_idx - 1, "):\n")
cat("Label:", df$Labels[example_idx + 1], "\n")
cat("\nSNPs found:", nrow(example_analysis$snps), "\n")
if (nrow(example_analysis$snps) > 0) {
  cat("First 5 SNPs:\n")
  for (i in 1:min(5, nrow(example_analysis$snps))) {
    snp <- example_analysis$snps[i, ]
    cat("  Position", snp$position, ":", snp$reference, "->", snp$alternate, "\n")
  }
}

cat("\nIndels found:", nrow(example_analysis$indels), "\n")
if (nrow(example_analysis$indels) > 0) {
  cat("First 3 Indels:\n")
  for (i in 1:min(3, nrow(example_analysis$indels))) {
    indel <- example_analysis$indels[i, ]
    cat("  Position", indel$position, ":", indel$type, ", Length:", indel$length, "\n")
  }
}

cat("\nStructural variants found:", nrow(example_analysis$structural_variants), "\n")
if (nrow(example_analysis$structural_variants) > 0) {
  cat("First 3 Structural Variants:\n")
  for (i in 1:min(3, nrow(example_analysis$structural_variants))) {
    sv <- example_analysis$structural_variants[i, ]
    cat("  Position", sv$position, ":", sv$type, ", Length:", sv$length, "\n")
  }
}
```
